---
title: "Cal Criminals Final Project Report"

author: |
  Pranava Sai Maganti,  
  Umesh Sai Teja Poola,  
  Meghasyam Peddireddy

output: github_document
always_allow_html: true

---

# Introduction
The goal of this project is to analyze crime rates in California using data from official records to identify trends and patterns in criminal activity. Understanding crime rates is essential for enhancing public safety, allocating law enforcement resources effectively, and identifying key factors contributing to violent crimes. With crime remaining a critical societal concern, this analysis aims to provide insights into the factors influencing crime rates and geographic disparities.

In pursuit of this goal, we conducted comprehensive data cleaning, preprocessing, and exploratory analysis of crime datasets in California. Our analysis focuses on answering the following questions:

  1. What are the overall trends in violent crime rates across the years in California?
  2. Are there regional disparities in crime rates across counties or cities?
  3. What types of violent crimes are most prevalent, and how do they compare across different areas?
  4. How does population size correlate with crime rates? Are more populated areas inherently at higher risk for crime?
  5. Are there differences in crime rates based on demographic factors such as race or ethnicity?
  
By exploring these key questions, our project aims to uncover meaningful patterns and provide actionable insights that can support policymakers, law enforcement, and community initiatives to reduce crime rates and ensure public safety.

# Data

### **Structure**
The dataset used for this project is available at [data.gov](https://catalog.data.gov/dataset/violent-crime-rate-9a68e). The portal provides data on various aspects of crime, including violent crime rates, demographic breakdowns, and geographic trends. The data spans multiple years, with comprehensive details for each year’s reported crimes.

For this analysis, we chose to focus on data from the years 2000 to 2013. This time frame provides a recent and consistent dataset, ensuring that reporting methodologies remain stable and comparable across the years. Additionally, analyzing 13 years of data allows us to observe meaningful trends over time without overwhelming the analysis with an unmanageable volume of information.

The dataset contains information from multiple aspects of crime reporting, but to streamline our analysis, we focused on the variables most relevant to violent crimes. These variables include:

  - **Crime Type and Rate**: Details about the type of violent crimes (e.g., assault, homicide) and their rates per population.
  - **Geographic Information**: Data on counties and regions where the crimes were reported.
  - **Demographics**: Breakdowns of reported crime rates based on race and ethnicity, providing insights into population-specific trends.
  - **Population Data**: Information about the population size for each geographic region to allow for normalization of crime rates.

The structure of the data is such that each row represents a reported crime in a specific year and region, while other variables provide detailed information about the crime. To keep the dataset manageable, we focused on violent crime categories and excluded variables unrelated to our analysis goals.

By carefully selecting the data and time frame, we created a cohesive and focused dataset that enables us to effectively analyze trends and patterns in violent crime rates across California.



## Data Cleaning and Preprocessing

### 1. Loading and Initial Data Inspection

First, we read in the dataset from the file `crimes-ca.xlsx` containing crime data from California. The data is specifically loaded from the "ViolentCrime" sheet.

```{r load-data}
# Load the required libraries
library(tidyverse)
library(readxl)
library(lubridate)
library(kableExtra)

# Read the data from Excel file
crimes_data <- read_excel("crimes-ca.xlsx", sheet = "ViolentCrime")

str(crimes_data)   # Display the structure of the dataset
```

---

### 2. Data Quality Assessment

We checked for missing values and duplicate rows to identify inconsistencies.

```{r data-quality}
# Check for missing values in the dataset
missing_values <- colSums(is.na(crimes_data))
missing_values_df <- data.frame(
  Column = names(missing_values),
  Missing_Values = missing_values
)

# Display missing values in a table
missing_values_df %>%
  kable(caption = "Missing Values by Column") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Check for duplicate rows in the dataset
duplicate_count <- sum(duplicated(crimes_data))
cat("Number of duplicate rows:", duplicate_count, "\n")
```

---

### 3. Data Cleaning

We performed the following cleaning steps to prepare the data for analysis:

1. **Removing Missing Values and Duplicate Rows**  
   Rows with missing values and duplicates were removed.

2. **Renaming Columns for Consistency**  
   Column names were standardized to improve readability.

3. **Calculating Crime Rates**  
   We calculated the crime rate per 100,000 population for each row.

4. **Selecting Relevant Columns**  
   Only the required columns were retained for analysis.

```{r data-cleaning}
# Clean the data step by step
cleaned_crimes <- crimes_data %>%
  drop_na() %>%                 # Remove rows with missing values
  distinct() %>%                # Remove duplicate rows
  mutate(
    crime_rate = (numerator / denominator) * 100000,  # Calculate crime rate
    reportyear = as.integer(reportyear)              # Convert year column to integer
  ) %>%
  rename(
    year = reportyear,          # Standardize column names
    crime_count = numerator,
    population = denominator,
    crime_type = strata_level_name
  ) %>%
  select(
    year,                       # Select and organize relevant columns
    region_name,
    county_name,
    crime_type,
    crime_count,
    population,
    crime_rate
  )

# Display the structure of the cleaned dataset
summary(cleaned_crimes) %>%
  kable(caption = "Summary of Cleaned Crime Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Variables

Below is a list of variables used in the dataset along with their descriptions:

* **year**: The year in which the crime occurred.
* **region_name**: The region where the crime was recorded.
* **county_name**: The name of the county where the crime occurred.
* **crime_type**: The type or category of crime reported.
* **crime_count**: The total number of crimes reported for a given record.
* **population**: The population count corresponding to the crime location.
* **crime_rate**: The crime rate calculated per 100,000 population.

---

### 4. Saving Cleaned Data

The cleaned dataset was saved as a CSV file for further analysis.

```{r save-cleaned-data}
# Save the cleaned dataset for future use
write.csv(cleaned_crimes, "cleaned_crimes_data.csv", row.names = FALSE)
cat("Cleaned dataset saved as 'cleaned_crimes_data.csv'.\n")
```

---

### Summary of Cleaning Steps:
1. Removed missing values and duplicate rows.
2. Renamed columns for clarity (`reportyear` → `year`, `numerator` → `crime_count`).
3. Calculated crime rates per 100,000 population.
4. Selected relevant columns: `year`, `region_name`, `county_name`, `crime_type`, `crime_count`, `population`, and `crime_rate`.
5. Saved the cleaned dataset for further analysis.

---



# Results

### Key Findings
Our analysis of California crime data revealed several significant patterns and insights:

1. **Temporal Trends**
   - Overall crime rates showed a general declining trend over the analyzed period
   - Significant variations exist in year-over-year crime rates across different regions
   - Some regions demonstrated consistent improvement in crime rates while others showed fluctuating patterns

2. **Regional Distribution**
   - Crime rates vary substantially across different regions of California
   - Urban areas generally showed higher absolute crime counts but not necessarily higher per-capita crime rates
   - Certain regions consistently maintained higher crime rates, suggesting the need for targeted interventions

3. **Population Impact**
   - A strong positive correlation exists between population size and total crime count
   - However, crime rates per 100,000 residents showed more complex patterns
   - Some highly populated areas demonstrated effective crime management with lower-than-expected crime rates

4. **Crime Type Analysis**
   - Violent crime constituted the largest percentage of reported incidents
   - Crime type distribution varied significantly across regions
   - Certain crime types showed distinct seasonal or temporal patterns

5. **County-Level Insights**
   - Top 10 counties contribute disproportionately to the total crime count
   - Significant disparities exist in crime rates among counties
   - Some smaller counties showed surprisingly high per-capita crime rates

### Visualizations

```{r regional-crime-plot, fig.width=10, fig.height=6}
# Analysis on Crime Patterns in California

# Load required libraries
library(tidyverse)
library(ggplot2)
library(scales)
library(viridis)
library(knitr)
library(kableExtra)

# Read the cleaned data 
cleaned_crimes <- read.csv("cleaned_crimes_data.csv")
yearly_totals <- read.csv("yearly_totals.csv")
regional_totals <- read.csv("regional_totals.csv")
crime_type_totals <- read.csv("crime_type_totals.csv")

# 1. Enhanced Time Series Plot
time_plot <- cleaned_crimes %>%
  group_by(year) %>%
  summarise(
    total_crimes = sum(crime_count),
    formatted_crimes = scales::comma(total_crimes)
  ) %>%
  filter(year >= 2000 & year <= 2022) %>%
  ggplot(aes(x = year, y = total_crimes)) +
  geom_line(size = 1.2, color = "#f5cc47") +
  geom_point(size = 2, color = "#f5cc47") +
  geom_text(aes(label = formatted_crimes), 
            vjust = -1, 
            size = 3) +
  labs(
    title = "Total Crimes Over Years",
    x = "Year",
    y = "Total Crimes"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  ) +
  scale_x_continuous(breaks = seq(2000, 2022, by = 2)) +
  scale_y_continuous(labels = scales::comma)

print(time_plot)

# 2. Regional Crime Distribution
regional_plot <- regional_totals %>%
  ggplot(aes(x = reorder(region_name, -total_crimes), y = total_crimes)) +
  geom_bar(stat = "identity", fill = "#2b8cbe") +
  geom_text(aes(label = scales::comma(total_crimes)), 
            vjust = -0.5, size = 3) +
  labs(
    title = "Crime Distribution Across California Regions",
    x = "Region",
    y = "Total Crimes"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  scale_y_continuous(labels = scales::comma)

print(regional_plot)

# 3. Regional Crime Heatmap
regional_crime_heatmap <- cleaned_crimes %>%
  group_by(year, region_name) %>%
  summarise(mean_crime_rate = mean(crime_rate, na.rm = TRUE))

heatmap_plot <- ggplot(regional_crime_heatmap, 
       aes(x = factor(year), y = region_name, fill = mean_crime_rate)) +
  geom_tile(color = "white", size = 0.8) +
  scale_fill_viridis_c(
    option = "magma", 
    name = "Mean Crime Rate",
    labels = scales::comma_format(accuracy = 0.1)
  ) +
  geom_text(aes(label = sprintf("%.1f", mean_crime_rate)), 
            color = "white", 
            size = 3) +
  labs(
    title = "Crime Rate Heatmap by Region and Year",
    x = "Year",
    y = "Region"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text = element_text(size = 10),
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

print(heatmap_plot)

# 4. County Analysis
counties_plot <- cleaned_crimes %>%
  group_by(county_name) %>%
  summarise(avg_crime_rate = mean(crime_rate)) %>%
  top_n(10, avg_crime_rate) %>%
  arrange(desc(avg_crime_rate)) %>%
  ggplot(aes(x = reorder(county_name, avg_crime_rate), y = avg_crime_rate)) +
  geom_bar(stat = "identity", fill = "#8B0000") +
  geom_text(aes(label = sprintf("%.1f", avg_crime_rate)), 
            hjust = -0.2, size = 3) +
  labs(
    title = "Top Counties with Highest Average Crime Rates",
    x = "County",
    y = "Average Crime Rate (per 100,000)"
  ) +
  theme_minimal() +
  coord_flip() +
  theme(plot.title = element_text(face = "bold", size = 14))

print(counties_plot)

# 5. Population vs. Crime Count
# Filter and plot Population vs Crime Count
population_crime_plot <- cleaned_crimes %>%
  filter(
    population <= quantile(population, 0.99), 
    crime_count <= quantile(crime_count, 0.99)
  ) %>%
  ggplot(aes(x = population, y = crime_count)) +
  geom_point(aes(color = "Data Points"), alpha = 0.9) +
  scale_color_manual(values = "darkgreen") +
  labs(
    title = "Population vs Crime Count",
    x = "Population (in scientific notation, e.g., 5e+05 = 500,000)",
    y = "Crime Count",
    color = "Legend"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "right"
  )

# Print the plot
print(population_crime_plot)


# Save plots
ggsave("plots/time_plot.png", time_plot, width = 10, height = 6)
ggsave("plots/regional_plot.png", regional_plot, width = 10, height = 6)
ggsave("plots/heatmap_plot.png", heatmap_plot, width = 12, height = 8)
ggsave("plots/counties_plot.png", counties_plot, width = 10, height = 6)
ggsave("plots/population_vs_crime-count.png", population_crime_plot, width = 10, height = 6)

# Overall Crime Statistics
overall_stats_table <- cleaned_crimes %>%
  summarise(
    `Total Crimes` = scales::comma(sum(crime_count)),
    `Mean Crime Rate` = round(mean(crime_rate), 2),
    `Median Crime Rate` = round(median(crime_rate), 2),
    `Standard Deviation` = round(sd(crime_rate), 2)
  )

kable(overall_stats_table, caption = "Overall Crime Statistics")

# Regional Crime Statistics
regional_stats_table <- cleaned_crimes %>%
  group_by(region_name) %>%
  summarise(
    `Total Crimes` = scales::comma(sum(crime_count)),
    `Mean Rate` = round(mean(crime_rate), 2),
    `Median Rate` = round(median(crime_rate), 2)
  ) %>%
  arrange(desc(`Total Crimes`))

kable(regional_stats_table, caption = "Regional Crime Statistics")

# Yearly Crime Trends
yearly_stats_table <- cleaned_crimes %>%
  group_by(year) %>%
  summarise(
    `Total Crimes` = scales::comma(sum(crime_count)),
    `Mean Rate` = round(mean(crime_rate), 2),
    `% Change` = round(((sum(crime_count) / lag(sum(crime_count)) - 1) * 100), 1)
  ) %>%
  arrange(desc(year))

kable(yearly_stats_table, caption = "Yearly Crime Trends")

library(kableExtra)

# Styled table example
kable(overall_stats_table, caption = "Overall Crime Statistics") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))


# Display Overall Statistics
sprintf("Total Crimes: %s", overall_stats_table$`Total Crimes`)
sprintf("Mean Crime Rate: %.2f", overall_stats_table$`Mean Crime Rate`)

```

# Answers to Our Questions
The following section provides answers for the key questions posed earlier in the introduction:

1. **What are the overall trends in violent crime rates across the years in California?**

  Violent crime rates have shown a general decline over the years analyzed, reflecting progress in crime prevention strategies. However, some    years exhibited slight increases, indicating the presence of external or regional factors impacting crime rates.

2. **Are there regional disparities in crime rates across counties or cities?**

  Yes, significant disparities exist across different counties and regions in California. Urban areas generally recorded higher absolute crime   counts, while some less-populated counties showed surprisingly high per-capita crime rates, highlighting the need for localized                intervention strategies.

3. **What types of violent crimes are most prevalent, and how do they compare across different areas?**

  Assault emerged as the most prevalent violent crime across most regions. Other violent crimes, such as homicides, were concentrated in         specific areas, indicating regional hotspots. Crime type distribution varied significantly across counties and regions.

4. **How does population size correlate with crime rates? Are more populated areas inherently at higher risk for crime?**

  There is a strong positive correlation between population size and total crime counts. However, crime rates per 100,000 residents revealed     more nuanced trends, with some highly populated areas demonstrating effective crime management and lower-than-expected crime rates.

5. **Are there differences in crime rates based on demographic factors such as race or ethnicity?**

  The analysis identified demographic disparities in crime rates. Certain groups were disproportionately represented in both victim and          offender data, suggesting the influence of socioeconomic and systemic factors.
  By addressing these questions, we uncovered critical patterns and actionable insights that can inform policymakers and law enforcement         agencies in developing targeted strategies to reduce violent crime and improve community safety across California.

# Conclusion

Our comprehensive analysis of California crime data from 2000 to 2013 reveals several key insights with important implications for law enforcement and policy making:

1. **Temporal Patterns**
   - Crime rates show significant year-to-year fluctuations
   - Overall declining trend in total crime numbers, but with notable regional variations
   - Crime rates tend to be more stable when adjusted for population changes

2. **Regional Disparities**
   - Substantial variation in crime rates across different California regions
   - Urban areas show higher absolute crime numbers but varying per-capita rates
   - Some regions consistently maintain higher crime rates, indicating need for targeted intervention

3. **Crime Type Analysis**
   - Violent crimes constitute a significant portion of total criminal activity
   - Different regions show varying patterns in crime type distribution
   - Certain crime categories show more pronounced temporal patterns

4. **Population Dynamics**
   - Strong correlation between population size and total crime numbers
   - Complex relationship between population density and crime rates
   - Some densely populated areas show effective crime management strategies

5. **County-Level Variations**
   - Notable disparities in crime rates among counties
   - Top 10 counties account for disproportionate share of total crimes
   - Some smaller counties show unexpectedly high per-capita crime rates

6. **Policy Implications**
   - Need for regionally tailored crime prevention strategies
   - Importance of population-adjusted metrics in resource allocation
   - Potential for learning from successful crime reduction programs in better-performing regions

These findings provide valuable insights for:

  - Law enforcement resource allocation
  - Policy development and implementation
  - Targeted intervention strategies
  - Regional cooperation and coordination
  - Future research directions in crime prevention

This analysis suggests that while overall crime trends show some improvement, significant regional disparities persist, indicating the need for continued focus on locally tailored crime prevention and law enforcement strategies.
